---
title: "U.S. Storm Event Analysis - 1985"
subtitle: "ANA 515 Assignment III"
author: "Anbazhagan Naresh"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
header-style: centered
output:
  html_document:
    theme:
      bootswatch: morph
---

<!-- Include Google Fonts -->
<link href="https://fonts.googleapis.com/css2?family=Montserrat:wght@400;700&display=swap" rel="stylesheet">

```{css, echo=FALSE}
h1.title {
  text-align: center;
  font-family: Montserrat;
  color: navy;
}
.h3 {
  text-align: left;
  font-family: Montserrat;
  color: navy;
}
.subtitle, .author, .date, .plot-title {
  text-align: left;
  font-family: Georgia;
  font-size: 20px;
  color: navy;
}
.subtitle {
  margin-top: 2em;
}
.author {
  margin-top: .5em;
}
.date {
  margin-top: .5em;
}
.body, p {
  text-align: justify;
  text-justify: inter-word;
  font-family: Montserrat;
  color: navy;
}

h1, h2{
  text-decoration: underline;
  text-decoration-color: yellow;
  text-underline-position: under;
  color: navy;
}

```


``` {r setup_packages, include = FALSE, warning=FALSE, message=FALSE}

# Specifying a CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# List of required packages
packages <- c("httr", "rvest", "dplyr", "stringr", "tidyr", "ggplot2", "lubridate", "knitr", "kableExtra", "DT")

# Function to check and install missing packages
check_packages <- function(pkgs){
  new_pkgs <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
  if(length(new_pkgs)) install.packages(new_pkgs, dependencies = TRUE)
}

# Call the function with the vector of packages
check_packages(packages)

# Load the packages with muted exception handling
sapply(packages, function(pkg) {
  if(!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    require(pkg, character.only = TRUE, quietly = TRUE)
  }
})

```


```{r download-source-data, echo=FALSE, message=FALSE, warning=FALSE}

# Defining the URL for the data directory
url <- "https://www1.ncdc.noaa.gov/pub/data/swdi/stormevents/csvfiles/"
local_file <- "StormEvents_details-ftp_v1.0_d1985.csv"

# Check if the file exists locally
if (!file.exists(local_file)) {
  # Scrape the web to find the most recent file for 1985
  page <- read_html(url)
  links <- page %>% html_nodes("a") %>% html_attr("href")
  file_link <- links[grepl("StormEvents_details-ftp_v1.0_d1985", links)]
  
  if (length(file_link) > 0) {
    # Get the most recent file
    download_link <- paste0(url, file_link[1])
    
    # Download the file
    download.file(download_link, destfile = local_file, method = "curl")
    message("File downloaded successfully.")
  } else {
    stop("No suitable file found on the website.")
  }
} else {
  message("Local file found and will be used.")
}
```


### Synopsis:

This analysis explores storm events in the United States for the year 1985 using data from NOAA's Storm Events Database. The dataset includes major weather-related storm events with details such as start and end dates, location, associated deaths, injuries, property damage, and other characteristics. The objective is to understand the distribution and characteristics of these events across different states, regions, and seasons. The raw dataset is linked to the below table:


```{r load-data, echo=FALSE, message=FALSE, warning=FALSE}
# Load the processed data
data <- read.csv(local_file)

# Check the column names
#colnames(data)

# Limiting the dataframe to the specified columns
filtered_columns <- c('BEGIN_YEARMONTH', 'EPISODE_ID', 'STATE', 'STATE_FIPS', 'CZ_NAME', 'CZ_TYPE', 'CZ_FIPS', 'EVENT_TYPE')
data <- data[filtered_columns]

# Arranging the data by state name
data <- data %>% arrange(STATE)

# Changing state and county names to title case
data$STATE <- str_to_title(data$STATE)
data$CZ_NAME <- str_to_title(data$CZ_NAME)

# Limiting to events listed by county FIPS (CZ_TYPE of 'C') and removing the CZ_TYPE column
data <- data %>% filter(CZ_TYPE == 'C') %>% select(-CZ_TYPE)

# Padding the state and county FIPS with '0' at the beginning and uniting the two columns to make one FIPS column
data$STATE_FIPS <- str_pad(data$STATE_FIPS, width = 3, side = "left", pad = "0")
data$CZ_FIPS <- str_pad(data$CZ_FIPS, width = 3, side = "left", pad = "0")
data$FIPS <- paste0(data$STATE_FIPS, data$CZ_FIPS)

# Changing all column names to lower case
colnames(data) <- tolower(colnames(data))

# Display the first few rows of the data
datatable(data, options = list(pageLength = 5, scrollX = TRUE))
```


The data was filtered to include only events listed by county FIPS, state and county names were converted to title case, and FIPS codes were padded to ensure consistency. The data was then merged with additional state information (area and region).

```{r merge-data, echo=FALSE, message=FALSE, warning=FALSE}
# Load the built-in state data
data("state")

# Create a dataframe with state name, area, and region
state_info <- data.frame(
  state = str_to_title(state.name),
  area = state.area,
  region = state.region
)

# Create a dataframe with the number of events per state
events_per_state <- data %>%
  group_by(state) %>%
  summarise(events = n())

# Merge the state information dataframe with the events per state dataframe
merged_data <- merge(events_per_state, state_info, by = "state")

# Remove any states that are not in the state information dataframe
merged_data <- merged_data %>%
  filter(!is.na(region))

# Display the merged data
kable(head(merged_data), format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

```{r save-data, echo=FALSE}
# Save the merged data for further inspection if needed
write.csv(merged_data, "merged_data.csv", row.names = FALSE)
```


### Summary table:

The summary table provides an overview of the total events, average events per month, and the area of each state.

```{r summarized-table, echo=FALSE}
summary_table <- merged_data %>%
  mutate(avg_events_per_month = events / 12) %>%
  select(state, events, avg_events_per_month, area, region)

# Display the summary table
kable(summary_table, col.names = c("State", "Total Events", "Avg Events per Month", "Area (sq miles)", "Region"), 
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Visualizations

### Bar Plot of Events per Region

The bar plot below shows the total number of storm events in 1985 by region. The Midwest and South regions have the highest number of events.

```{r bar-plot, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}
ggplot(merged_data, aes(x = region, y = events, fill = region)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Total Number of Storm Events in 1985 by Region",
    x = "Region",
    y = "Number of Events"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


### Box Plot of Events per Region

The box plot below displays the distribution of storm events in 1985 by region. The South region shows a wide range of event counts with some outliers, indicating variability in storm activity.

```{r box-plot-of-events-region, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}
ggplot(merged_data, aes(x = region, y = events, fill = region)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Storm Events in 1985 by Region",
    x = "Region",
    y = "Number of Events"
  ) +
  theme_minimal()
```


### Histogram of Event Distribution

The histogram below shows the distribution of the number of events in 1985. The majority of states experienced fewer than 500 events, with some states experiencing significantly higher numbers.

```{r histogram-of-event-dist, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}
ggplot(merged_data, aes(x = events, fill = region)) +
  geom_histogram(binwidth = 10, alpha = 0.7, position = "identity") +
  labs(
    title = "Histogram of Storm Events in 1985",
    x = "Number of Events",
    y = "Frequency"
  ) +
  theme_minimal()
```


### Scatter Plot of Events by State Area

The scatter plot below visualizes the number of events by state area. There is no clear relationship between the size of the state and the number of storm events.

```{r scatter-plot-by-area, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}
ggplot(merged_data, aes(x = area, y = events, color = region)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Scatter Plot of Events by State Area",
    x = "State Area (sq miles)",
    y = "Number of Events"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, color = "slateblue"))
```


### Heatmap of Events by State and Month

The heatmap below shows the number of storm events by state and month in 1985. This visualization helps to identify patterns and peak periods of storm activity for each state.

```{r heatmap-by-month, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}

# Adding year and month columns for the heatmap
data$year <- as.integer(substr(data$begin_yearmonth, 1, 4))
data$month <- as.integer(substr(data$begin_yearmonth, 5, 6))

events_by_month <- data %>%
  group_by(state, month) %>%
  summarise(events = n(), .groups = 'drop')

ggplot(events_by_month, aes(x = factor(month), y = state, fill = events)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(
    title = "Heatmap of Events by State and Month in 1985",
    x = "Month",
    y = "State",
    fill = "Number of Events"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, color = "slateblue"))

```

### Pie Chart of Event Types


```{r event-chart, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}
event_type_counts <- data %>%
  group_by(event_type) %>%
  summarise(events = n()) %>%
  arrange(desc(events))

event_type_counts$fraction <- event_type_counts$events / sum(event_type_counts$events)
event_type_counts$ymax <- cumsum(event_type_counts$fraction)
event_type_counts$ymin <- c(0, head(event_type_counts$ymax, n=-1))

ggplot(event_type_counts, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=event_type)) +
  geom_rect() +
  coord_polar(theta="y") +
  xlim(c(2, 4)) +
  labs(
    title = "Pie Chart of Event Types",
    fill = "Event Type"
  ) +
  theme_void() +
  theme(plot.title = element_text(size = 14, color = "slateblue"), legend.position = "right")
```


### Seasonal Patterns

The analysis of seasonal patterns helps to understand how storm events are distributed throughout the year.

##### Defining Seasons
To analyze the seasonal patterns, the data was divided into four seasons:

Winter: January, February, December

Spring: March, April, May

Summer: June, July, August

Fall: September, October, November

The bar plot below shows the number of storm events in 1985 for each season. Summer had the highest number of storm events, followed by Spring.


```{r season-chart, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}
# Add month column
data$month <- as.integer(substr(data$begin_yearmonth, 5, 6))

# Define seasons
data <- data %>%
  mutate(season = case_when(
    month %in% c(12, 1, 2) ~ "Winter",
    month %in% c(3, 4, 5) ~ "Spring",
    month %in% c(6, 7, 8) ~ "Summer",
    month %in% c(9, 10, 11) ~ "Fall"
  ))

# Summarize data by season
events_by_season <- data %>%
  group_by(season) %>%
  summarise(events = n()) %>%
  arrange(factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

# Bar plot of events by season
ggplot(events_by_season, aes(x = season, y = events, fill = season)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Storm Events by Season in 1985",
    x = "Season",
    y = "Number of Events"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, color = "slateblue"), legend.position = "none")
```


### Line Plot of Events Over Time

The line plot below shows the number of storm events each month in 1985. The peak is observed in June, indicating increased storm activity during early summer.

```{r line-plot-time-series, echo=FALSE, warning=FALSE, fig.width=10, fig.height=6}

events_by_month <- data %>%
  group_by(month) %>%
  summarise(events = n())

# Print the data to verify
# kable(events_by_month, format = "html") %>%
#  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

ggplot(events_by_month, aes(x = month, y = events)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Line Plot of Events Over Time by Month",
    x = "Month",
    y = "Number of Events"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, color = "slateblue"))

```

